---
title: "Untitled"
author: "Nikolas Santamaria"
date: "2024-04-03"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)

list.of.packages <- c("pls","dplyr","reshape2","here","plotrix","ggplot2","gridExtra")
invisible(lapply(list.of.packages, library, character.only = TRUE))

df <- read.csv('/Users/niks/Desktop/USDA/Spring 2024/Data/2023 NxIrr_hyperspectral_txt/8-8-2023.csv')
df2 <- read.csv('/Users/niks/Desktop/USDA/Spring 2024/Data/2023 NxIrr_hyperspectral_txt/8-24-2023.csv')



data_2023 <- list(
 "08/08/2023" = df,
 "08/24/2023" = df2
)



plot_numbers <- list(
  '4114' = 1, '1111' = 2, '5115' = 3, '3113' = 4, '2112' = 5,
  '6116' = 6, '3123' = 7, '1121' = 8, '6126' = 9, '5125' = 10,
  '4124' = 11, '2122' = 12, '6226' = 13, '1221' = 14, '5225' = 15,
  '4224' = 16, '3223' = 17, '2222' = 18, '4214' = 19, '5215' = 20,
  '3213' = 21, '6216' = 22, '1211' = 23, '2212' = 24, '3323' = 25,
  '6326' = 26, '4324' = 27, '5325' = 28, '2322' = 29, '1321' = 30,
  '4314' = 31, '1311' = 32, '2312' = 33, '3313' = 34, '6316' = 35,
  '5315' = 36
)

```


```{r}


for (key in names(data_2023)) {
  value <- data_2023[[key]] # Access the value by key
  value <- value %>% rename(Wavelength = X) %>%
  mutate(Wavelength = as.integer(Wavelength)) # Convert 'Wavelength' to integer

  #convert wavelength column to int
  data_2023[[key]] <- value
}


```



```{r pressure, echo=FALSE}
# Define the treatment_100 vector
irrigation_100 = c(1:6, 19:22, 31:36)

# Define the full sequence
full_seq = 1:36

# Find elements in full_seq not in treatment_100
irrigation_70 = full_seq[!full_seq %in% irrigation_100]

```







Function to sort based on first row and convert to long


```{r}

leaf_sorter_long <- function(df, date){
  # Identify columns for top_leaf and middle_leaf based on first row
  # Assuming 'df' is your dataframe in R and the first column is 'Wavelength'
  
  # First column name (presumed 'Wavelength' here)
  wavelength_column <- names(df)[1]
  # Initialize vectors to hold column names for each category
  top_leaf_columns <- c(wavelength_column)  # Include 'Wavelength' column
  middle_leaf_columns <- c(wavelength_column)  # Include 'Wavelength' column
  
  # Iterate over the column names, starting from the second column
  for (col in names(df)[-1]) {  # Exclude the first column during iteration
      if (grepl("TL", df[1, col])) {
          # If "TL" is found in the first row of this column, add to top_leaf_columns
          top_leaf_columns <- c(top_leaf_columns, col)
      } else if (grepl("ML", df[1, col])) {
          # If "ML" is found, add to middle_leaf_columns
          middle_leaf_columns <- c(middle_leaf_columns, col)
      }
  }
  
  # Subset the original dataframe to create top_leaf and middle_leaf dataframes
  top_leaf <- df[, top_leaf_columns]
  middle_leaf <- df[, middle_leaf_columns]
  
  # Optionally, remove the first row from both dataframes as it's used for categorization
  top_leaf <- top_leaf[-1, ]
  middle_leaf <- middle_leaf[-1, ]
  
  

  
  top_long <- pivot_longer(top_leaf, cols = -Wavelength, names_to = "Measurement", values_to = "Value")
  top_long['Leaf_Location'] <- "Top"
  middle_long <- pivot_longer(middle_leaf, cols = -Wavelength, names_to = "Measurement", values_to = "Value")
  middle_long['Leaf_Location'] <- "Middle"
  
  top_middle_long <- rbind(top_long, middle_long)
  top_middle_long$Date <- date

  top_middle_long
}



```



```{r}
library(dplyr)
library(purrr)

#running for each day

df_long <- data_2023 %>% 
  map2(names(.), ~leaf_sorter_long(.x, .y)) %>%
  bind_rows()

```


```{r}

add_plot_number <- function(df, plot_numbers) {
  # Assuming df_long should be df as per the function argument
  df <- df %>%
    mutate(
      Plot = sapply(Measurement, function(x) {
        key <- substr(x, 2, 5)  # Extract the key from the Measurement
        if (key %in% names(plot_numbers)) {
          return(plot_numbers[[key]])  # Return the corresponding plot number
        } else {
          return(NA)  # Return NA if the key is not found
        }
      }),
      Treatment = sapply(Measurement, function(x) {
        key <- substr(x, 2, 5)  # Reuse the extracted key from Measurement
        if(key %in% names(plot_numbers)) {
          # Extract and return the first digit of the key
          return(substr(key, 1, 1))
        } else {
          return(NA)  # Return NA if the key is not found
        }
      })
    )
  
  df
}

df_long <- add_plot_number(df_long, plot_numbers)
```




```{r}
library(dplyr)

# Assuming df_long is your dataframe

# Define plots with 100% irrigation treatment
irrigation_100 <- c(1:6, 19:22, 31:36)

# Define the full sequence of plot numbers
full_seq <- 1:36

# Define plots with 70% irrigation treatment by excluding those in irrigation_100 from full_seq
irrigation_70 <- full_seq[!full_seq %in% irrigation_100]

# Assuming your dataframe df_long has a column named Plot
df_long <- df_long %>%
  mutate(Irrigation = ifelse(Plot %in% irrigation_100, "100%", 
                             ifelse(Plot %in% irrigation_70, "70%", NA)))

```


```{r}

biomass_csv <- read.csv('/Users/niks/Desktop/USDA/Spring 2024/Data/2023_biomass.csv')
biomass <- biomass_csv[,c("Date", "Plot", "Carbon.per.Plant..g.", "Nitrogen.per.Plant..g.")]
```


```{r}

biomass$Date <- as.Date(biomass$Date, format="%m/%d/%y")
df_long$Date <- as.Date(df_long$Date, format="%m/%d/%Y")
biomass <- biomass %>% rename(
  Carbon ="Carbon.per.Plant..g.", 
  Nitrogen = "Nitrogen.per.Plant..g."
)
biomass <- biomass %>%
  mutate(Date = if_else(Date == as.Date("2023-08-09"), as.Date("2023-08-08"), Date)) %>%
  mutate(Date = if_else(Date == as.Date("2023-09-27"), as.Date("2023-08-24"), Date))


df_long <- df_long %>%
  left_join(biomass, by = c("Plot", "Date"))


```

```{r}

### Chosing the combinations of groupings to perform PLS

# Define unique combinations of Day, Irrigation, Treatment
#This is for daily Irrigation/Treatment combos


```{r}
df_wide <- df_long %>% 
  pivot_wider(names_from = Wavelength, values_from = Value)

```

```{r}
`%notin%` <- Negate(`%in%`)


Start.wave <- 500
End.wave <- 2400
wv <- seq(Start.wave,End.wave,1)
Spectra <- as.matrix(df_wide[,names(df_wide) %in% wv])
colnames(Spectra) <- c(paste0("Wave_",wv))
sample_info <- df_wide[,names(df_wide) %notin% seq(350,2500,1)]

```


```{r}
#Create PLSR data
plsr_data <- data.frame(sample_info, Spectra)

```
### Step 6: Data Already Cleaned

### Step 7: Create Cal/Val datasets
```{r}
method <- "dplyr" #base/dplyr
# base R - a bit slow
# dplyr - much faster

split_data <- spectratrait::create_data_split(dataset=plsr_data, approach=method, 
                                              split_seed=7529075, prop=0.8, 
                                              group_variables=c("Irrigation", "Treatment"))

names(split_data)
```

```{r}
cal.plsr.data <- split_data$cal_data
val.plsr.data <- split_data$val_data


rm(split_data)

# Datasets:
print(paste("Cal observations: ",dim(cal.plsr.data)[1],sep=""))
print(paste("Val observations: ",dim(val.plsr.data)[1],sep=""))

```

```{r}


inVar = "Nitrogen"

make_cal_val_plots <- function(inVar){
  text_loc <- c(max(hist(cal.plsr.data[,paste0(inVar)], plot=FALSE)$counts),
                max(hist(cal.plsr.data[,paste0(inVar)], plot=FALSE)$mids))
  cal_hist_plot <- qplot(cal.plsr.data[,paste0(inVar)],geom="histogram",
                         main = paste0("Calibration Histogram for ",inVar),
                         xlab = paste0(inVar),ylab = "Count",fill=I("grey50"),col=I("black"),
                         alpha=I(.7)) + 
    annotate("text", x=text_loc[2], y=text_loc[1], label= "1.",size=10)
  val_hist_plot <- qplot(val.plsr.data[,paste0(inVar)],geom="histogram",
                         main = paste0("Validation Histogram for ",inVar),
                         xlab = paste0(inVar),ylab = "Count",fill=I("grey50"),col=I("black"),
                         alpha=I(.7))
  histograms <- grid.arrange(cal_hist_plot, val_hist_plot, ncol=2)
  
  ggsave(filename = file.path(paste0(inVar,"_Cal_Val_Histograms.png")), 
       plot = histograms, device="png", width = 30, height = 12, units = "cm", 
       dpi = 300)
# output cal/val data
write.csv(cal.plsr.data,file=file.path(paste0(inVar,'_Cal_PLSR_Dataset.csv')),
          row.names=FALSE)
write.csv(val.plsr.data,file=file.path(paste0(inVar,'_Val_PLSR_Dataset.csv')),
          row.names=FALSE)
}

make_cal_val_plots("Nitrogen")
make_cal_val_plots("Carbon")

```



Step 8. Create calibration and validation PLSR datasets

```{r}
### Format PLSR data for model fitting 
cal_spec <- as.matrix(cal.plsr.data[, which(names(cal.plsr.data) %in% 
                                              paste0("Wave_",wv))])
cal.plsr.data <- data.frame(cal.plsr.data[, which(names(cal.plsr.data) %notin% 
                                                    paste0("Wave_",wv))], 
                            Spectra=I(cal_spec))
head(cal.plsr.data)[1:5]
```


```{r}

val_spec <- as.matrix(val.plsr.data[, which(names(val.plsr.data) %in% 
                                              paste0("Wave_",wv))])
val.plsr.data <- data.frame(val.plsr.data[, which(names(val.plsr.data) %notin% 
                                                    paste0("Wave_",wv))],
                            Spectra=I(val_spec))
head(val.plsr.data)[1:5]
```
```{r}
cal.plsr.data$Spectra <- apply(cal.plsr.data$Spectra, 2, as.numeric)  # Convert each column to numeric
val.plsr.data$Spectra <- apply(val.plsr.data$Spectra, 2, as.numeric)  # Convert each column to numeric



par(mfrow=c(1,2)) # B, L, T, R
spectratrait::f.plot.spec(Z=cal.plsr.data$Spectra,wv=wv,
            plot_label="Calibration")
text(550,95,labels = "2.",cex=3)
spectratrait::f.plot.spec(Z=val.plsr.data$Spectra,wv=wv,
            plot_label="Validation")
```

```{r}
### Use permutation to determine the optimal number of components
if(grepl("Windows", sessionInfo()$running)){
  pls.options(parallel = NULL)
} else {
  pls.options(parallel = parallel::detectCores()-1)
}

method <- "firstMin" #pls, firstPlateau, firstMin
random_seed <- 7529075
seg <- 80
maxComps <- 16
iterations <- 50
prop <- 0.70
if (method=="pls") {
  nComps <- spectratrait::find_optimal_components(dataset=cal.plsr.data, targetVariable=inVar,
                                                  method=method, 
                                                  maxComps=maxComps, seg=seg, 
                                                  random_seed=random_seed)
  print(paste0("*** Optimal number of components: ", nComps))
} else {
  nComps <- spectratrait::find_optimal_components(dataset=cal.plsr.data, targetVariable=inVar,
                                                  method=method, 
                                                  maxComps=maxComps, iterations=iterations, 
                                                  seg=seg, prop=prop, 
                                                  random_seed=random_seed)
}

```

```{r}
### Fit final model - using leave-one-out cross validation
plsr.out <- plsr(as.formula(paste(inVar,"~","Spectra")),scale=FALSE,ncomp=nComps,
                 validation="LOO",trace=FALSE,data=cal.plsr.data)
fit <- plsr.out$fitted.values[,1,nComps]
pls.options(parallel = NULL)

# External validation fit stats
text_loc <- c(max(RMSEP(plsr.out, newdata = val.plsr.data)$comps),
              RMSEP(plsr.out, newdata = val.plsr.data)$val[1])
par(mfrow=c(1,2)) # B, L, T, R
pls::RMSEP(plsr.out, newdata = val.plsr.data)
```

```{r}
plot(pls::RMSEP(plsr.out,estimate=c("test"),newdata = val.plsr.data), main="MODEL RMSEP",
     xlab="Number of Components",ylab="Model Validation RMSEP",lty=1,col="black",cex=1.5,lwd=2)
text(text_loc[1],text_loc[2],labels = "4.", cex=2)
box(lwd=2.2)

pls::R2(plsr.out, newdata = val.plsr.data)
```

```{r}
plot(pls::R2(plsr.out,estimate=c("test"),newdata = val.plsr.data), main="MODEL R2",
     xlab="Number of Components",ylab="Model Validation R2",lty=1,col="black",cex=1.5,lwd=2)
box(lwd=2.2)
```
```{r}
#calibration
cal.plsr.output <- data.frame(cal.plsr.data[, which(names(cal.plsr.data) %notin% 
                                                      "Spectra")],
                              PLSR_Predicted=fit,
                              PLSR_CV_Predicted=as.vector(plsr.out$validation$pred[,,
                                                                                   nComps]))
cal.plsr.output <- cal.plsr.output %>%
  mutate(PLSR_CV_Residuals = PLSR_CV_Predicted-get(inVar))
head(cal.plsr.output)
```

```{r}
cal.R2 <- round(pls::R2(plsr.out,intercept=F)[[1]][nComps],2)
cal.RMSEP <- round(sqrt(mean(cal.plsr.output$PLSR_CV_Residuals^2)),2)

val.plsr.output <- data.frame(val.plsr.data[, which(names(val.plsr.data) %notin% 
                                                      "Spectra")],
                              PLSR_Predicted=as.vector(predict(plsr.out, 
                                                               newdata = val.plsr.data, 
                                                               ncomp=nComps, 
                                                               type="response")[,,1]))
val.plsr.output <- val.plsr.output %>%
  mutate(PLSR_Residuals = PLSR_Predicted-get(inVar))
head(val.plsr.output)
```
```{r}

val.R2 <- round(pls::R2(plsr.out,newdata=val.plsr.data,intercept=F)[[1]][nComps],2)
val.RMSEP <- round(sqrt(mean(val.plsr.output$PLSR_Residuals^2)),2)

rng_quant <- quantile(cal.plsr.output[,inVar], probs = c(0.001, 0.999))
cal_scatter_plot <- ggplot(cal.plsr.output, aes(x=PLSR_CV_Predicted, y=get(inVar))) + 
  theme_bw() + geom_point() + geom_abline(intercept = 0, slope = 1, color="dark grey", 
                                          linetype="dashed", size=1.5) + 
  xlim(rng_quant[1], rng_quant[2]) + 
  ylim(rng_quant[1], rng_quant[2]) +
  labs(x=paste0("Predicted ", paste(inVar), " (units)"),
       y=paste0("Observed ", paste(inVar), " (units)"),
       title=paste0("Calibration: ", paste0("Rsq = ", cal.R2), "; ", 
                    paste0("RMSEP = ", cal.RMSEP))) +
  theme(axis.text=element_text(size=18), legend.position="none",
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5)) + 
  annotate("text", x=rng_quant[1], y=rng_quant[2], label= "5.",size=10)

cal_resid_histogram <- ggplot(cal.plsr.output, aes(x=PLSR_CV_Residuals)) +
  geom_histogram(alpha=.5, position="identity") + 
  geom_vline(xintercept = 0, color="black", 
             linetype="dashed", size=1) + theme_bw() + 
  theme(axis.text=element_text(size=18), legend.position="none",
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))

rng_quant <- quantile(val.plsr.output[,inVar], probs = c(0.001, 0.999))
val_scatter_plot <- ggplot(val.plsr.output, aes(x=PLSR_Predicted, y=get(inVar))) + 
  theme_bw() + geom_point() + geom_abline(intercept = 0, slope = 1, color="dark grey", 
                                          linetype="dashed", size=1.5) + 
  xlim(rng_quant[1], rng_quant[2]) + 
  ylim(rng_quant[1], rng_quant[2]) +
  labs(x=paste0("Predicted ", paste(inVar), " (units)"),
       y=paste0("Observed ", paste(inVar), " (units)"),
       title=paste0("Validation: ", paste0("Rsq = ", val.R2), "; ", 
                    paste0("RMSEP = ", val.RMSEP))) +
  theme(axis.text=element_text(size=18), legend.position="none",
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))

val_resid_histogram <- ggplot(val.plsr.output, aes(x=PLSR_Residuals)) +
  geom_histogram(alpha=.5, position="identity") + 
  geom_vline(xintercept = 0, color="black", 
             linetype="dashed", size=1) + theme_bw() + 
  theme(axis.text=element_text(size=18), legend.position="none",
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))

# plot cal/val side-by-side
scatterplots <- grid.arrange(cal_scatter_plot, val_scatter_plot, cal_resid_histogram, 
                             val_resid_histogram, nrow=2, ncol=2)

```